Ensemble Methods for Robust Feature Selection

### 2016-08-05 paper/paper.tex
typos

Feature selection is a method used in machine learning to find a small subset of features in order to build a more accurate, simpler and faster model. It is often used with high dimensional data. Moreover it allows domain experts to gain a better understanding of the data, by focusing on the selected subset. However, in domains such as biomedical applications subsequent analysis are costly. Therefore not only the model performance but also the robustness of the feature selection method is important. Domain experts would prefer a more stable algorithm to have more confidence in the selected features. One approach for more robust results are ensemble methods.

@@

Symmetrical Uncertainty (SU) \citep{press1996numerical} is an entropy based measurement. It is the normalized mutual information between the feature and the class. The mutual information measures the influence of the knowledge of the class on the feature value in terms of entropy and normalizes it. It is defined as follows:
\begin{equation}
  \label{eq:su}
  SU(F,C) = 2 \frac{H(F) - H(F|C)}{H(F) + H(C)} \textrm{, where } H(\cdot) \textrm{ is the entropy.}
\end{equation}

More details about mutual information can be found in \cite{paninski2003estimation}. The second feature selection is RELIEFF \citep{kononenko1997overcoming}. For each sample $x_i$ the k nearest neighboring samples, using the $L_1$ norm, belonging the same class $\{\textrm{Near-hit}_{i, j}\}_{j=1..k}$ and the ones of the opposite class $\{\textrm{Near-miss}_{i, j}\}_{j=1..k}$ are selected. Then the difference is taken to update the weight vector representing how well each feature distinguishes samples of the two classes, as can be seen in equation \ref{eq:relief}. In all the experiments k is set to 5, which, empirically, gives satisfactory results.
\begin{equation}
  \label{eq:relief}
  W_i = W_i - \frac{1}{k}\sum_{j=1}^{k}|x_i - \textrm{Near-hit}_{i,j}| + \frac{1}{k}\sum_{j=1}^{k} |x_i - \textrm{Near-miss}_{i,j}|
\end{equation}

@@

The last feature selection method used is Lasso logistic regression \citep{wheeler2010lasso}. It is a logistic regression with a $L_1$ norm as penalty for the weight vector. For the hyperparameter $C$ a grid search with $10$ different parameters in logarithmic scale between $10^{-5}$ and $10^3$ is done. The parameters are evaluated with a stratified 3-fold cross-validation.

\begin{equation}
    \min_{w,c} \quad \|w\|_1 + C\sum_{i=1}^n \log(\exp(y_i(x_i^Tw + c))+1)
  \end{equation}
  The advantage of this method is that we obtain a sparse weight vector as solution. This means only a small subset of features will have a significant high weight. In this work Lasso logistic regression is also called Lasso.
  \section{Ensemble methods}
  The different feature selection methods give as output a weight vector representing how well each feature distinguishes samples of the two classes. The ensemble method used aggregates the different weights vectors. Since different methods do not always have the same weight range, they are normalized to the range $[0,1]$. Moreover for the methods which can only rank data, such as methods based on RFE, their ranking are considered as weights, and so subsequently normalized the same way. 
  Those normalized weights are combined using different techniques. The methods used in this paper can be grouped in three distinct categories:

\section{Ensemble methods}

The different feature selection methods give as output a weight vector representing how well each feature distinguishes samples of the two classes. Those are then be combined by the ensemble method. Since different methods do not always have the same weight range, they are normalized to the range $[0,1]$. Moreover for the methods which can only rank data, such as methods based on RFE, their ranking are considered as weights, and so subsequently normalized the same way.

Those normalized weights are combined using different schemes. The methods used in this paper can be grouped in two distinct categories

@@

The features of the artificial dataset is generated with different Gaussian distributions, depending on whether they are used in the labeling process or not. It consists of $100$ significant features and $9900$ fake features. The significant features and fake features both consists of correlated and uncorrelated features. The correlation between the correlated features is randomly generated. The ratio of correlated features and uncorrelated features are for the significant and the fake features $0.5$. 
\begin{align}
f \sim \mathcal{N}(0, 1) + \mathcal{N}(0, \Sigma)  \quad \text{with} \quad \Sigma_{ij} \sim \mathcal{U}(0,1)
\end{align}
Therefore the artificial consists of $50$ correlated significant features, $50$ uncorrelated significant features, $4950$ correlated fake features and $4950$ uncorrelated fake features. We will call the uncorrelated features univariate features and the correlated features multivariate features in this work.
% \text{unsignificant} \quad & f \sim 0.5 * \left[ N(0, 1) + U(-1,1) \right]
To generate balanced binary labels, the significant features are first summed, then the median of the sums is subtracted, finally the sign is taken.

@@

To measure the robustness, 10 equally sized subsamples are drawn from the dataset. The feature selector runs on all of them, and the best ones are taken. The subsets are then compared with a similarity measure. Given that the model performance is measured on the best 1\% of the features, the order of the feature should not be of importance but only if one is or is not part of the best subset. The Jaccard Index is used because only how many features are common to both sets is important for this measure. For two subsets of features $F_i$ and $F_j$ it is defined as:

@@

\subsection{Results}
In this section we present our results for the experiments. %TODO what results
\subsubsection{Distribution of feature selectors}
\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{images/feature_weights_hist_arcene_artificial.png}
  \caption{The histograms of the feature weights for all used feature selectors on the arcene and artificial dataset.}
  \label{fig:feature_weights_hist_arcene_artificial}
\end{figure}

@@

  \caption{The features weights once for only the significant features and once for all features. In figure \ref{fig:feature_weights_plot_artificial_only_features} in the white area are the significant multivariate features and in the diagonal hatched area are the significant univariate features. In figure \ref{fig:feature_weights_plot_artificial_all} in the diagonal hatched area are all significant features, in the white area are the fake multivariate features and in the horizontal hatched area are the fake univariate features. }
  \label{fig:feature_weights_plot_artificial}
\end{figure}
The weights of the features can be seen in figure \ref{fig:feature_weights_plot_artificial}.
All feature selectors achieved similar results for the artificial dataset except for SU. SU ranks all features with a zero weight, making it not better than a random feature selection on this dataset. On the contrary the other feature selectors recognize parts of the inherent structure of the data. The univariate features are in general ranked lower than the multivariate features. There seems to be no difference in weighting between the significant und fake univariate features. However the significant multivariate features are ranked significantly higher than the fake multivariate features. As a consequence only multivariate features are selected. The reason why the feature selectors weight the significant univariate features this low is that the univariate features do not contribute significantly to the classification. If we use only the significant multivariate features with the same classification function with which the labels are created, we achieve an accuracy of $0.98$. On the other hand using the significant univariate features we only achieve an accuracy of $0.52$. Therefore a feature which correlates with other significant features has significantly more impact on the classification of a data point than a uniform feature.

@@

To compare the different feature selection methods in relation of robustness and accuracy, the single feature selection and the three ensemble methods achieving the best results are plotted with accuracy against stability. For the stability the Jaccard index with $1\%$ of the features is used. For the accuracy the all classifiers described in \ref{sec:experiments}  reused. For each feature selection method the standard deviation for the accuracy and the stability on the different subsets are plotted. The results for the different classifiers are similar. In figure \ref{fig:boxplot_svm} the results for the C-SVM classifier can be observed. The other results can be seen in the appendix. %TODO boxplots for plots
As we can see in figure \ref{fig:boxplot_svm} the ensemble methods achieve good results on each dataset. While single feature selection methods achieve better results in terms of accuracy and robustness on some datasets, they fail on others. However the ensemble methods keep their performance for each dataset. On the arcene dataset they even achive better results than any single feature selector.

@@

\section{Discussion}
We have shown that ensemble methods can keep up with results of single feature selectors and are even superior to the tested feature selectors in terms of generality over different datasets. Further knowledge about the weakness and strength of certain feature selectors can exploited to create good ensembles of feature selectors. For the future it has to be tested if the ensemble methods in this work show similar performance on other datasets.

@@

\appendix
\label{appendix}
\section{Python libraries}
\label{apd:libraries}
For the feature selection methods SU and RELIEFF we used the scikit-feature library from \cite{Li-etal16}.
For C-SVM, RFE, Lasso logistic regression we used the scikit-sklearn library from \cite{scikit-learn}
\section{Additional plots}
\label{apd:plots}
\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{images/feature_weights_hist.png}
  \caption{The histogram of the feature weights for all feature selectors used for all datasets used.}
  \label{fig:feature_weights_hist}
\end{figure}

>>>>

Feature selection is a method used in machine learning to find a small subset of features to build a more accurate, simpler and faster model. It is often used with high dimensional data. Moreover it allows domain experts to gain a better understanding of the data, by focusing on the selected subset. However, in domains such as biomedical applications subsequent analysis are costly. Thus not only the model performance but also the robustness of the feature selection method is important. Domain experts would prefer a more stable algorithm to have more confidence in the selected features. One approach for more robust results are ensemble methods.

@@

Symmetrical Uncertainty (SU) \citep{press1996numerical} is an entropy based measurement. It is the normalized mutual information between the feature and the class. The mutual information measures the influence of the knowledge of the class on the feature value in terms of entropy and normalizes it. It is defined as follows:
\begin{equation}
  \label{eq:su}
  SU(F,C) = 2 \frac{H(F) - H(F|C)}{H(F) + H(C)} \textrm{, where } H(\cdot) \textrm{ is the entropy.}
\end{equation}
More details about mutual information can be found in \cite{paninski2003estimation}.

The second feature selection is RELIEFF \citep{kononenko1997overcoming}. For each sample $x_i$ the k nearest neighboring samples, using the $L_1$ norm, belonging the same class $\{\textrm{Near-hit}_{i, j}\}_{j=1..k}$ and the ones of the opposite class $\{\textrm{Near-miss}_{i, j}\}_{j=1..k}$ are selected. Then the difference is taken to update the weight vector representing how well each feature distinguishes samples of the two classes, as can be seen in equation \ref{eq:relief}. In all the experiments k is set to 5, which, empirically, gives satisfactory results.
\begin{equation}
  \label{eq:relief}
  W_i = W_i - \frac{1}{k}\sum_{j=1}^{k}|x_i - \textrm{Near-hit}_{i,j}| + \frac{1}{k}\sum_{j=1}^{k} |x_i - \textrm{Near-miss}_{i,j}|
\end{equation}

@@

The last feature selection method used is Lasso logistic regression \citep{wheeler2010lasso}. It is a logistic regression with a $L_1$ norm as penalty for the weight vector. The hyperparameter $C$ is estimated with a grid search with $10$ different parameters in logarithmic scale between $10^{-5}$ and $10^3$. The parameters are evaluated with a stratified 3-fold cross-validation.

\begin{equation}
    \min_{w,c} \quad \|w\|_1 + C\sum_{i=1}^n \log(\exp(y_i(x_i^Tw + c))+1)
  \end{equation}
  The advantage of this method is to obtain a sparse weight vector as solution. Hence only a small subset of features have a significant high weight. In this work Lasso logistic regression is also called Lasso for simplicity.

\section{Ensemble methods}
 The ensemble method used in this paper combine different feature selection methods. Those give as output a weight vector representing how well each feature distinguishes samples of the two classes. Thus the ensemble method has to aggregate the different weights vectors. Since different methods do not always have the same weight range, they are normalized to the range $[0,1]$. Moreover for the methods which can only rank data, such as methods based on RFE, their ranking are considered as weights, and so subsequently normalized the same way.

 The normalized weights are combined using different techniques. The ones used in this paper can be grouped in two distinct categories:

@@

The features of the artificial dataset is generated with different Gaussian distributions, depending on whether they are used in the labeling process or not. It consists of $100$ significant features and $9900$ fake features. The significant features and fake features both consists of correlated and uncorrelated features. The correlation between the correlated features is randomly generated. The ratio of correlated features and uncorrelated features are for the significant and the fake features $0.5$. 
\begin{align}
f \sim \mathcal{N}(0, 1) + \mathcal{N}(0, \Sigma)  \quad \text{with} \quad \Sigma_{ij} \sim \mathcal{U}(0,1)
\end{align}
Therefore the artificial consists of $50$ correlated significant features, $50$ uncorrelated significant features, $4950$ correlated fake features and $4950$ uncorrelated fake features. In this work, the uncorrelated features and the correlated features are also respectively called univariate and  multivariate features.
To generate balanced binary labels the significant features are first summed, then the median of the sum is subtracted, finally the sign is taken.

@@

To measure the robustness, 10 equally sized subsamples are drawn from the dataset. The feature selector runs on all of them, and the best ones are taken. The subsets are then compared with a similarity measure. Given that the model performance is measured on the best 1\% of the features, the order of the feature should not be of importance but only if one is or is not part of the best subset. The Jaccard Index is used because only how many features are common to both sets is important. For two subsets of features $F_i$ and $F_j$ it is defined as:

@@

\subsection{Results}
In this section the results of the different experiments are presented. First the distribution of the weights given by the feature selectors is analyzed. Secondly the results on the artificial dataset is discussed. Then the influence of different combination of feature selectors is analyzed. Finally the a comparison between the ensemble methods and the feature selectors is made.

\subsubsection{Distribution of feature selectors}
\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{images/feature_weights_hist_arcene_artificial.png}
  \caption{The histograms of the feature weights for all used feature selectors on the arcene and artificial dataset.}
  \label{fig:feature_weights_hist_arcene_artificial}
\end{figure}

@@

  \caption{The features weights once for only the significant features and once for all features. In figure \ref{fig:feature_weights_plot_artificial_only_features} in the white area are the significant multivariate features and in the diagonal hatched area are the significant univariate features. In figure \ref{fig:feature_weights_plot_artificial_all} in the diagonal hatched area are all significant features, in the white area are the fake multivariate features and in the horizontal hatched area are the fake univariate features. }
  \label{fig:feature_weights_plot_artificial}
\end{figure}
The weights of the features can be seen in figure \ref{fig:feature_weights_plot_artificial}.
All feature selectors achieve similar results for the artificial dataset except for SU. SU ranks all features with a zero weight, making it not better than a random feature selection on this dataset. On the contrary the other feature selectors recognize parts of the inherent structure of the data. The univariate features are in general ranked lower than the multivariate features. There seems to be no difference in weighting between the significant und fake univariate features. However the significant multivariate features are ranked significantly higher than the fake multivariate features. As a consequence only multivariate features are selected. The reason why the feature selectors weight the significant univariate features this low is that the univariate features do not contribute significantly to the classification. If only the significant multivariate features are used with classification function used to generate the labels, an accuracy of $0.98$ is obtained. On the other hand using only the significant univariate features the accuracy drops to $0.52$. Therefore a feature which correlates with other significant features has significantly more impact on the classification of a data point than a uniform feature.

@@

To compare the different feature selection methods in relation of robustness and accuracy, the single feature selection and the three ensemble methods achieving the best results are plotted with accuracy against stability. For the stability the Jaccard index with $1\%$ of the features is used. For the accuracy the all classifiers described in \ref{sec:experiments}  reused. For each feature selection method the standard deviation for the accuracy and the stability on the different subsets are plotted. The results for the different classifiers are similar. In figure \ref{fig:boxplot_svm} the results for the C-SVM classifier can be observed. The other results can be seen in the appendix. %TODO boxplots for plots
As can be seen in figure \ref{fig:boxplot_svm} the ensemble methods achieve good results on each dataset. While single feature selection methods achieve better results in terms of accuracy and robustness on some datasets, they fail on others. However the ensemble methods keep their performance for each dataset. On the arcene dataset they even achieve better results than any single feature selector.

@@

\section{Discussion}
This work shows that ensemble methods can keep up with results of single feature selectors and are even superior to the tested feature selectors in terms of generality over different datasets. Further knowledge about the weakness and strength of certain feature selectors can exploited to create good combinations of feature selectors. For the future it has to be tested if the ensemble methods in this work show similar performance on other datasets.

@@

\appendix
\label{appendix}
\section{Python libraries}
\label{apd:libraries}
For the feature selection methods SU and RELIEFF come from the scikit-feature library (\cite{Li-etal16}).
For C-SVM, RFE, Lasso logistic regression come from the scikit-sklearn library (\cite{scikit-learn}).
\section{Additional plots}
\label{apd:plots}
\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{images/feature_weights_hist.png}
  \caption{The histogram of the feature weights for all feature selectors used for all datasets used.}
  \label{fig:feature_weights_hist}
\end{figure}
